Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-16 paper.tex
minor update text

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux from of clusters: Fornax
  (upper left), Coma (upper right), Virgo (lower left), and Perseus
  (lower right). We show CR induced emission (red solid), a leptophilic
  (LP) model that includes both final state radiation and IC upscattered
  CMB, dust and starlight (green dash-dotted), and the benchmark $\Kp$
  model (BM) that includes continuum emission, and IC upscattered CMB, dust
  and starlight (blue dashed). The arrows show the differential upper
  limits set by Fermi-LAT in the energy ranges $0.2-1$~GeV,
  $1-10$~GeV, and $10-100$~GeV from left to right. We show the
  individual components for the BM model and LP model for Virgo and
  Perseus, respectively. The flux from the clusters is integrated out
  to $\rvir$ using a point spread function of $0.1\deg$ and includes
  the boost from substructures. We find that the lower GeV energy
  regime is most constraining due to the peak sensitivity of Fermi at
  these energies and use this regime to get upper limits on boost
  factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

In Fig.~\ref{fig15}, we compare Fermi-LAT flux upper limits to the
flux predictions by an analytical CR model \cite{2010MNRAS.409..449P}
for CR induced gamma-rays. For consistency, we use beta-model fits to
ROSAT data of X-ray bright clusters for our gamma-ray flux estimates
as an input \cite{2002ApJ...567..716R}). We derive the central gas
densities for each cluster by following the approach outlined in
\cite{1999ApJ...517..627M} which assumes that the ICM is
isothermal. The central gas density is found by combining X-ray
temperature with X-ray luminosity, X-ray detection radius, density
core radius and density slope for each cluster. These predictions are
superior to those that use luminosity-mass scaling relation to infer
the expected gamma-ray flux---a method used by reference
\cite{2010ApJ...717L..71A} to compare to Fermi limits.  In this
analytic model, we employ the fact that CRs accelerated at structure
formation shocks and transported over cosmic time. This leads to a CR
distribution with a universal spectral form and an almost universal
spatial distribution. Depending on whether or not we account for the
bias of ``artificial galaxies'' in cosmological simulations, we derive
an optimistic or conservative limit of the expected gamma-ray emission
from decaying pions (bracketing our ignorance of the contribution of
cluster galaxies to the overall gamma-ray flux from a cluster). Note
that both models assume the maximally allowed acceleration efficiency
for strong shocks that transfers 50\% of the shock-dissipated energy
(kinetic energy corrected for adiabatic compressional heating at the
shock) to CRs. These efficiencies drop quickly for weaker shocks
\cite{2007A&A...473...41E} which dominate the gravitational energy
dissipation inside galaxy clusters \cite{2006MNRAS.367..113}. These
efficiencies will have to be corrected down if this value is not
realized at structure formation shocks. Secondly, these simulations
neglect active CR transport such as streaming and diffusion relative
to the gas, i.e. we assume that advective transport is the dominant
one and CRs are tightly coupled to the gas via flux-frozen magnetic
fields tangled on sufficiently small scales. Evidence from the
statistics of radio halos suggests that this might not be the case for
relaxed clusters, arguing that CR streaming in those objects could be
responsible for a decrease in the central CR number density. This
effect would also reduce the expected gamma-ray flux by up to a factor
of five \cite{2011A&A...527A..99E}. We note that the tightest Fermi
limits are obtained in the 1-10 GeV regime due to the highest
sensitivity of Fermi-LAT there. While the effective area of Fermi-LAT
increases up to these energies, the typical photon spectra decrease as
a function of energy: combining those effects selects this energy
range to be most sensitive.  In this energy band, the Fermi limits
come close to the flux predictions of Virgo, Coma and Norma. Using
integrated flux limits from Fermi, we arrive at very similar results.
We emphasize that by using our analytic modeling, the slight
discrepancy of the scaling relation-based prediction with the Fermi
limit on Norma \cite{2010ApJ...717L..71A} is resolved: Norma simply
seems to be less bright than an average cluster of the same mass as
Norma. The insensitive limit on Perseus comes about since the central
AGN in NGC1275 has turned on over the last years, making it hard to
determine a comparable limit as obtained for the other clusters.

>>>>

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters: Fornax
  (upper left), Coma (upper right), Virgo (lower left), and Perseus
  (lower right). We show CR induced emission (red solid), a leptophilic
  (LP) model that includes both final state radiation and IC upscattered
  CMB, dust and starlight (green dash-dotted), and the benchmark $\Kp$
  model (BM) that includes continuum emission, and IC upscattered CMB, dust
  and starlight (blue dashed). The arrows show the differential upper
  limits set by Fermi-LAT in the energy ranges $0.2-1$~GeV,
  $1-10$~GeV, and $10-100$~GeV from left to right. We show the
  individual components for the BM model and LP model for Virgo and
  Perseus, respectively. The flux from the clusters is integrated out
  to $\rvir$ using a point spread function of $0.1\deg$ and includes
  the boost from substructures. We find that the lower GeV energy
  regime is most constraining due to the peak sensitivity of Fermi at
  these energies and use this regime to get upper limits on boost
  factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

In Fig.~\ref{fig15}, we compare Fermi-LAT flux upper limits to the
flux predictions by an analytical CR model \cite{2010MNRAS.409..449P}
for CR induced gamma-rays. For consistency, we use beta-model fits to
ROSAT data of X-ray bright clusters for our gamma-ray flux estimates
as an input \cite{2002ApJ...567..716R}). We derive the central gas
densities for each cluster by following the approach outlined in
\cite{1999ApJ...517..627M} which assumes that the ICM is
isothermal. The central gas density is found by combining X-ray
temperature with X-ray luminosity, X-ray detection radius, density
core radius and density slope for each cluster. These predictions are
superior to those that use luminosity-mass scaling relation to infer
the expected gamma-ray flux---a method used by reference
\cite{2010ApJ...717L..71A} to compare to Fermi limits.  In this
analytic model, we employ the fact that CRs accelerated at structure
formation shocks and transported over cosmic time. This leads to a CR
distribution with a universal spectral form and an almost universal
spatial distribution. Depending on whether or not we account for the
bias of ``artificial galaxies'' in cosmological simulations, we derive
an optimistic or conservative limit of the expected gamma-ray emission
from decaying pions (bracketing our ignorance of the contribution of
cluster galaxies to the overall gamma-ray flux from a cluster). Note
that both models assume the maximally allowed acceleration efficiency
for strong shocks that transfers 50\% of the shock-dissipated energy
(kinetic energy corrected for adiabatic compressional heating at the
shock) to CRs. These efficiencies drop quickly for weaker shocks
\cite{2007A&A...473...41E} which dominate the gravitational energy
dissipation inside galaxy clusters \cite{2006MNRAS.367..113P}. These
efficiencies will have to be corrected down if this value is not
realized at structure formation shocks. Secondly, these simulations
neglect active CR transport such as streaming and diffusion relative
to the gas, i.e. we assume that advective transport is the dominant
one and CRs are tightly coupled to the gas via flux-frozen magnetic
fields tangled on sufficiently small scales. Evidence from the
statistics of radio halos suggests that this might not be the case for
relaxed clusters, arguing that CR streaming in those objects could be
responsible for a decrease in the central CR number density. This
effect would also reduce the expected gamma-ray flux by up to a factor
of five \cite{2011A&A...527A..99E}. We note that the tightest Fermi
limits are obtained in the 1-10 GeV regime due to the highest
sensitivity of Fermi-LAT there. While the effective area of Fermi-LAT
increases up to these energies, the typical photon spectra decrease as
a function of energy: combining those effects selects this energy
range to be most sensitive.  In this energy band, the Fermi limits
come close to the flux predictions of Virgo, Coma and Norma. Using
integrated flux limits from Fermi, we arrive at very similar results.
We emphasize that by using our analytic modeling, the slight
discrepancy of the scaling relation-based prediction with the Fermi
limit on Norma \cite{2010ApJ...717L..71A} is resolved: Norma simply
seems to be less bright than an average cluster of the same mass as
Norma. The insensitive limit on Perseus comes about since the central
AGN in NGC1275 has turned on over the last years, making it hard to
determine a comparable limit as obtained for the other clusters.
